Convergence rate of MAP estimates \\ for the exponential family

### 2021-10-29 aistats.tex
Derive gamma bound for natural MAP

In higher dimensions, $X \sim \cN(0, \mu_*)$, $X\in\real^d$, $T(X) = XX^\top$, and the mean parameter $\mu_*$ 
is a $d \times d$ symmetric, positive definite covariance matrix with $p = \frac{d(d+1)}{2}$ degrees of freedom.
Note that here $d$ denotes the dimensionality of the data $X$, rather than the dimensionality of the parameters $\mu_*$.

@@

\subsection{Bounding the expected natural parameters for the MAP}
Before proving a convergence rate for the MAP, we need to bound the expectation of its inverse, hence the following lemma.
We introduce the notation $(z)_+ = \max(0,z)$.
\begin{lemma}[Expected MAP natural parameter]\label{lem:expected-map-natural-parameter-gaussian}
	Define the variable $a = n_0 \frac{\mu_0}{\mu^*}$, which characterizes 
	the importance of the prior relative to the true parameters.
	The expectation of the natural parameter of the MAP of $\cN(0,\mu_*)$ is bounded by
\begin{equation}
		\frac{n_0+n}{a+n}
		=\frac{\mu^*}{\mu_n}
		\leq \expect{\frac{\mu^*}{\MAPm}} 
		= \expect{\frac{\MAPt}{\nat^*}} 
		\leq
		\frac{n_0 +n}{a+(n-2)_+} \; , \forall n\geq 0 \; .
		\label{eq:lemma_nat_bound}
	\end{equation}
\end{lemma}

\begin{proof}
	The lower bound can be readily obtained by applying Jensen's inequality 
	to the convex function $x\mapsto\inv{x}$ for $x>0$.
	The upper bound requires more work.
	To start, let us plug in the definition of $\MAPm$
\alignn{
		\expect{\frac{\mu^*}{\MAPm}} 
		= \expect{\frac{\MAPt}{\nat^*}} 
		= \expect{\frac{(n_0+n) \mu^*}{n_0 \mu_0 + \sum_i X_i^2}}  
		= (n_0 + n) \expect{\inv{n_0 \frac{\mu_0}{\mu^*}+ \sum_i \frac{X_i^2}{ \mu^*} }} 
		= (n_0 +n) \expect{\inv{a + \chi^2(n)}} \; ,
}
	where $\chi^2(n) = \sum_i \frac{X_i^2}{ \mu^*}$ is a chi-square random variable of degree $n$ and $a=n_0 \frac{\mu_0}{\mu^*}$. The expectation reads
\alignn{
 	\expect{\inv{a + \chi^2(n)}}  
	= \int_0^\infty \inv{a+x} \frac{x^{\frac{n}{2}-1}e^{-\half[x]} }{2^{\half[n]}\Gamma(\half[n])} dx \; .
}
	This $1/(a+x)$ is hard to deal with.
	Fortunately, there is a trick to express it as an integral itself, with an exponential that fits well with the $\chi^2$-density,
\alignn{
		&\int_1^\infty e^{-\frac{zt}{2}} dt = \frac{2}{z} e^{-\half[z]} \\
		\implies &\inv{z} = \half e^{\half[z]} \int_1^\infty e^{-\frac{zt}{2}} dt \\
		\implies & \boxed{\inv{a+x} = \half e^{\frac{a+x}{2}} \int_1^\infty e^{- \frac{(a+x)t}{2}}dt  \; .}
} 
We can then plug this integral into the expectation and apply Fubini's theorem to get a tractable integral on the inside
\alignn{
		2^{\half[n]}\Gamma(\half[n]) \expect{\inv{a + \chi^2(n)}}  
		&= \int_0^\infty \frac{x^{\frac{n}{2}-1}e^{-\half[x]} }{a+x} dx \\
		&= \int_0^\infty dx\  x^{\frac{n}{2}-1}e^{-\half[x]} \half e^{\half[a+x]} \int_1^\infty  dt\  e^{-\half[(a+x)t]}  \\
		&= \half e^{\half[a]} \int_1^\infty dt \ e^{-\half[a t]}   \int_0^\infty dx \  x^{\frac{n}{2}-1} e^{-\half[x t]}  \; ,
}
	assuming we can exchange the order of integration.
	Now with the change of variable $x = 2\frac{y}{t}$, eg $dx = 2\frac{dy}{t}$, the inner integral becomes
	\begin{align}
		\expect{\inv{a + \chi^2(n)}}  
		&= \frac{e^{\half[a]} }{2^{\half[n]+1}\Gamma(\half[n]) }\int_1^\infty dt \ e^{-\half[a t]}   \int_0^\infty 2\frac{dy}{t} \  (2\frac{y}{t})^{\frac{n}{2}-1} e^{-y} \\
		&= \frac{e^{\half[a]} }{2\Gamma(\half[n]) } 
		\underbrace{\int_1^\infty  \frac{e^{-\half[a t]}}{t^{\frac{n}{2}}}  dt}_{E_{\half[n]}(\half[a])  } \ 
		\underbrace{\int_0^\infty  y^{\frac{n}{2}-1} e^{-y} dy}_{\Gamma(\half[n])}
	\end{align}
	so we finally get  the formula valid for all $n\geq 1$
	\begin{align}
		\boxed{\expect{\inv{a + \chi^2(n)}}  = \half e^{\frac{a}{2}} E_{\half[n]}(\half[a])}
		\label{eq:hard_expectation}
	\end{align}
	where the {generalized exponential integral function} is defined as 
	\begin{align}
		E_k(z) = \int_1^\infty \frac{e^{-z t} }{t^k} dt \; .
	\end{align}
	Now our goal is to bound this function with simpler functions.
	Fortunately, mathematicians have been working on these integrals for decades.
	For $n \geq 2$, eg $k\geq 1$, we have \href{https://dlmf.nist.gov/8.19.E21}{the general bound} \citep[Eq.~8.19.21]{DLMF}
	\begin{align}
		e^x E_k(x) 
		&\leq \inv{x + k - 1} \\
		\iff \expect{\inv{a + \chi^2(n)}}
		&\leq \inv{a + n - 2} \\
		\iff \expect{\frac{\mu^*}{\MAPm}} 
		&\leq \frac{n +n_0}{a + n - 2}.
	\end{align}
We are left with a special case when  $n=1$, eg $k=\half$. 
Then we can use the trivial bound 
\begin{align}
	\expect{\inv{a + X^2}} < \inv{a}.
	\label{eq:trivial_bound}
\end{align}
to conclude the proof.
\end{proof}

>>>>

For the sake of simplicity, in higher dimension we focus on Gaussian covariance estimation and avoid the general Wishart discussion. 
In higher dimensions, $X \sim \cN(0, \mu_*)$, $X\in\real^d$, $T(X) = XX^\top$, and the mean parameter $\mu_*$ 
is a $d \times d$ symmetric, positive definite covariance matrix with $p = \frac{d(d+1)}{2}$ degrees of freedom.
Note that here $d$ denotes the dimensionality of the data $X$, rather than the dimensionality of the parameters $\mu_*$.

@@

\subsection{Bounding the expected natural parameters for the MAP}
Before proving a convergence rate for the MAP, we need to bound the expectation of its inverse, hence the following lemma.
We introduce the notation $(z)_+ = \max(0,z)$.
\begin{lemma}[Expected MAP natural parameter]\label{lem:expected-map-natural-parameter-gaussian}
	Define the variable $a = n_0 \frac{\mu_0}{\mu^*}$, which characterizes 
	the importance of the prior relative to the true parameters.
	The expectation of the natural parameter of a MAP of $\Gamma(\alpha,\beta)$ is bounded by
\begin{equation}
		\frac{n_0+n}{a+n}
		=\frac{\mu^*}{\mu_n}
		\leq \expect{\frac{\mu^*}{\MAPm}} 
		= \expect{\frac{\MAPt}{\nat^*}} 
		\leq
		\frac{n_0 +n}{a+(n-\inv{\alpha} )_+} \; , \forall n\geq 0 \; .
		\label{eq:lemma_nat_bound}
	\end{equation}
\end{lemma}

\begin{proof}
	The lower bound can be readily obtained by applying Jensen's inequality 
	to the convex function $x\mapsto\inv{x}$ for $x>0$.
	The upper bound requires more work.
	To start, let us plug in the definition of $\MAPm$
\alignn{
		\expect{\frac{\mu^*}{\MAPm}} 
		= \expect{\frac{\MAPt}{\nat^*}} 
		= \expect{\frac{(n_0+n) \mu^*}{n_0 \mu_0 + \sum_i X_i}}  
		= \expect{\frac{n_0+n}{n_0 \frac{\mu_0}{\mu^*}+ \sum_i \frac{X_i}{ \mu^*} }} 
		= \expect{\frac{n_0+n}{a + \Gamma(n\alpha, \alpha)}} \; ,
}
	where  $\sum_i \frac{X_i^2}{ \mu^*} \sim \Gamma(n\alpha, \alpha)$ is a Gamma random variable and $a=n_0 \frac{\mu_0}{\mu^*}$. 
	Further note that 
\alignn{
	\expect{\frac{n_0+n}{a + \Gamma(n\alpha, \alpha)}} 
	= \frac{n_0 + n}{a} \expect{\frac{1}{1 + \Gamma(n \alpha , a \alpha) }} \; .
}
This kind of integrals can be expressed with {generalized exponential integral functions}
	\begin{align}
		E_k(z) = \int_1^\infty \frac{e^{-z t} }{t^k} dt \; ,
	\end{align}
with \href{http://dlmf.nist.gov/8.19.E4}{the formula} \citep[Eq.~8.19.4]{DLMF}
\alignn{
	\expect{\frac{1}{1 + \Gamma(\alpha , \beta) }}
	= \frac{\beta^{\alpha}}{\Gamma(\alpha)} \int_0^\infty \frac{x^{\alpha-1}e^{-\beta x} }{1+x}  dx 
	= \beta e^{\beta} E_\alpha (\beta) \; .
}
Overall we get
\alignn{
\expect{\frac{\mu^*}{\MAPm}}  = (n_0 +n) \alpha e^{a \alpha} E_{n\alpha} ( a \alpha) 
}
	Now our goal is to bound this generalized exponential integral with simpler functions.
	Fortunately, mathematicians have been working on these integrals for decades.
	For instance , we have \href{https://dlmf.nist.gov/8.19.E21}{the general bound} \citep[Eq.~8.19.21]{DLMF}
	\begin{align}
		e^x E_k(x) 
		&\leq \inv{x + k - 1} , \ \forall k>1\\
		\iff \expect{\frac{\mu^*}{\MAPm}} 
		&\leq \frac{(n_0 +n)\alpha}{a\alpha + n\alpha - 1}
		= \frac{n_0 +n}{a + n - \inv{\alpha}} , \ \forall n >\inv{\alpha} \; .
	\end{align}
We are left with a special case when  $n\leq \alpha$. 
Then we can use the trivial bound 
\begin{align}
	\expect{\inv{a + X^2}} < \inv{a}.
	\label{eq:trivial_bound}
\end{align}
to conclude the proof.
\end{proof}

For small $n$. a more subtle alternative uses \href{http://dlmf.nist.gov/8.19.E12}{the recurrence relationship} \cite[Eq.~8.19.12]{DLMF}
\alignn{
	\alpha E_{\alpha+1}(\beta) + \beta E_\alpha(\beta) 
	= e^{-\beta} 
}
and combines it with the inequality \citep[Eq.~8.19.21]{DLMF} to get
\alignn{
	\beta e^\beta E_\alpha(\beta)
 	= 1 - \alpha e^\beta E_{\alpha+1}(\beta)
 	\leq 1 - \frac{\alpha}{\alpha + 1 + \beta}
 	 = \frac{\beta + 1}{\alpha + \beta + 1} \; .
}
Back to the MAP, we get
\alignn{
	\expect{\frac{\mu^*}{\MAPm}} 
	\leq \frac{n_0 + n}{a} \frac{a\alpha + 1}{n \alpha + a \alpha +1}
	= \frac{n_0 + n}{a + n +\inv{\alpha}} \paren{1 + \inv{a \alpha}}
}
